---
title: 'NGLY1 exploration: part 2'
output:
  html_document:
    df_print: paged
---

# Background
Previously, I looked into a single general query defined by Nuria: how are NGLY1 and AQP1 gene related to each other, within 3 links of each other? Based on that exploration, I identified the following things to change:

1. remove duplicate relationships; add clutter and it's unclear what the directionality of the relationship is.
2. group "redundant" relationships together; combine orthologues and encoding relationships
3. categorize types of connections between the terms into semantic groups.
4. break apart graph network and make more like a parallel coordinates plot, ideally w/ small multiples
5. create hierarchical filtering: rather than have a single graph, have series of vis to categorize, help whittle down the results.

# These data
... are based on a more specific, targeted query between keys terms in the NGLY1 pathway. They all have the relationship:
[source] == [source orthologue] --> [disorder] (physical manifestation / phenotype) --> [gene] == [human orthologue of gene] --> [physiological pathway] --> [target]

# Import + cleanup
```{r start, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(data.table)
library(llamar)

data_dir = '~/GitHub/ngly1-exploration/data/input/'
q1 = read_csv(paste0(data_dir, '2015-01-15_ngly1-engase.csv'))



node_edge_cols = function(n_nodes){
  nodes = paste0('node', 1:n_nodes)
  edges = paste0('edge', 1:(n_nodes-1))
  
  q1 %>% 
  separate(path, sep = '\\},\\{', into = node_edge_cols(4), split = ",")
  
  # interleave the nodes and edges together
  ne = paste(nodes, edges, sep = ',', collapse = ',')
  
  # convert to a list
  ne = str_split(ne, pattern = ',', simplify = TRUE)
  
  # remove the last element-- it's a leftover from the fact that nodes/edges are uneven, so R will repeat the first edge.
  ne = ne[1:(length(ne) - 1)]
  
  return(ne)
}


cleanup_neo4j = function(df) {
  
}
```
```{r}
q2 = jsonlite::fromJSON(txt=paste0(data_dir, '2018-01-15_NFE2L1-apq1.json'), flatten = TRUE) 

q2 = q2$data[[1]]

convert_graph = function(data) {
  # interwoven list of nodes and connections
  row = data %>% data.frame(stringsAsFactors = FALSE)
  
  num_entries =  nrow(row)
  wide = row  %>% 
    mutate(node_num = ceiling(1:num_entries/2),
           NE = str_c(c(rep(c('node', 'edge'), floor(num_entries/2)), 'node'), node_num),
           name = ifelse(NE %like% 'node', preflabel,
                         ifelse(NE %like% 'edge', property_label, NA)))
  
  wide = wide %>% 
    select(NE, name) %>% 
    spread(NE, name)
  
  return(wide)
}

long_graph = function(data) {
  # interwoven list of nodes and connections
  row = data %>% data.frame(stringsAsFactors = FALSE)
  
  num_entries =  nrow(row)
  
  long = row  %>% 
    mutate(node_num = ceiling(1:num_entries/2),
           NE = c(rep(c('node', 'edge'), floor(num_entries/2)), 'node'),
           name = ifelse(NE %like% 'node', preflabel,
                         ifelse(NE %like% 'edge', property_label, NA)),
                      url = na_if(reference_uri, '<NA>'))
  
  long = long %>% 
    select(node_num, NE, name, url)
  
  return(long)
}

q2_t = lapply(q2, convert_graph) %>% bind_rows() 
q2_t = q2_t %>% mutate(path_num = 1:nrow(q2_t))

q2_long = lapply(q2, long_graph) %>% bind_rows()  %>% 
  mutate(path_num = rep(1:length(q2), each = 13)) 

q2_edges = q2_long %>% 
  filter(NE == 'node') %>% 
  group_by(path_num) %>% 
  arrange(node_num, path_num) %>% 
  # transform into source/targets
  mutate(source = name,
         target = lead(name)) %>% 
  filter(!is.na(target))

# descriptive stats
n_nodes = q2_long %>% filter(NE == 'node') %>% group_by(node_num) %>% 
  count(name) %>% 
  arrange(node_num, desc(n)) %>% 
  ungroup() %>% 
  select(-node_num) %>% 
  rename(value = n)
kable(n_nodes)

n_edges = q2_edges %>% 
  ungroup() %>% 
  group_by(node_num) %>% 
  count(source, target) %>% 
  arrange(node_num, desc(n)) %>% 
    ungroup() %>% 
  select(-node_num) %>% 
  mutate(value = 1)
kable(n_edges)

# double checking that every edge is the same
q2_long %>% filter(NE == 'edge') %>% group_by(node_num) %>% count(name) %>% arrange(desc(n)) %>% kable()
```
```{r export}
write_csv(n_edges, '~/GitHub/ngly1-exploration/data/output/demo_edges.csv')
write_csv(n_nodes, '~/GitHub/ngly1-exploration/data/output/demo_nodes.csv')
x = NULL
x$nodes = n_nodes
x$links = n_edges
jsonlite::write_json(x, '~/GitHub/ngly1-exploration/src/data/demo_graph.json')
```


```{r, plot_nodes}
ggplot(n_nodes, aes(y = value, x = forcats::fct_reorder(name, value))) + 
  geom_bar(stat = 'identity') + 
  coord_flip() +
  facet_wrap(~node_num, scales = 'free', nrow = 1) +
  theme_ylab() +
  theme(panel.spacing = unit(1, 'lines'))

```


